Reconstructing contact network parameters from viral phylogenies

### 2016-07-01 doc/paper/changes.tex
updates

Second, we examined the possible Monte Carlo error by increasing the resolution
of the ABC approximation (more particles, more simulated datasets per particle,
and more iterations). The results were nearly identical, TODO

@@

  TODO

@@

We struggled with the phrasing of this particular point and are happy to have
recieved some guidance. TODO

>>>>

Second, we examined the possible Monte Carlo error by increasing the resolution
of the ABC approximation (more particles, more simulated datasets per particle,
and more iterations). The results were similar; surprisingly, increased
settings made the estimation about 10\% worse. We suspect this may be simply
due to stochastic effects. In any event, this result demonstrates that Monte
Carlo error from the SMC algorithm is likely not responsible for the observed
estimation errors.

@@

We have substantially expanded the discussion of the synthetic data results,
including this point about $m$. The weaker identifiability in lower $\alpha$
values is related to the degree distributions of the networks, and the upward
biases in $I$ and $N$ are discussed in relation to the epidemic trajectory of
the SI model. Our intuition about $m$ is weaker, but we include this point by
the reviewer and hypothesize that $m$ may be more readily estimable from the
speed of epidemic expansion rather than the shape of the transmission tree.

@@

We struggled with the phrasing of this particular point and are happy to have
recieved some guidance. The statement now reads as follows: ``If $I$ is small
relative to $N$, very few of the infected nodes will gain any new neighbours.
Thus, the outcome of a second simulation on the same network will likely be
very similar.''

### 2016-07-01 doc/paper/paper.Rnw
updates

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
    \caption[
        One-dimensional marginal posterior distributions of BA model
        parameters estimated by \textit{netabc} from a simulated transmission
        tree.
    ]{
        One-dimensional marginal posterior distributions of BA model parameters
        estimated by \textit{netabc} from a simulated transmission tree. Dashed
        lines indicate true values, solid lines indicate posterior means, and
        shaded areas show 95\% highest posterior density intervals.
    }
    \label{fig:abc1}
\end{figure}

@@

    joint.I <- d[is.na(fixed), abs(I-true_I)]
    marginal.I <- d[fixed == "N+alpha+m", abs(I-true_I)]
@

\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
of each of the BA parameters on the same simulated transmission tree. The
estimates of $m$ are apparently unaffected by marginalizing out the other
parameters, corroborating the previous experiments' findings that $m$ is not an
identifiable parameter from scaled tree shapes. Compared to allowing all
parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
    \Sexpr{ round((joint.alpha - marginal.alpha) * 100 / joint.alpha) }\%,
    \Sexpr{ round((joint.I - marginal.I) * 100 / joint.I) }\%, and
    \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
when all other parameters were fixed. Neither the accuracy nor the precision of
any parameter estimates appeared to improve when more computing power was used
for SMC (\cref{fig:moresmc}).}

@@

% OLD STUFF
 Since the
prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
N$), there is more prior mass on high $N$ values.

This can be
illustrated by imagining that we add a small number of nodes to a network after
the epidemic simulation has already been completed. It is possible that none of
these new nodes attains a connection to any infected node. Thus, running the
simulation again on the new, larger network could produce the exact same
transmission tree as before. We note also that \del{our accurate estimates of
$I$ may have been influenced by} this prior\del{, which} places more mass on
low $I$ values. However, the estimate of $I$ was very high for the HET/Botswana
data, suggesting that a strong enough signal in the data can
overcome the prior. \add{Furthermore, when $I$ was estimated marginally with
fixed $N$, the accuracy of the estimate improved even though there was no
longer any extra prior mass on low $I$ values.}

% END OLD STUFF

\add{The number of nodes in the network, $N$, exhibited the most variation in
terms of its effect on tree shape. There was almost no measurable difference
between trees simulated under different $N$ values when the number of infected
nodes $I$ was small (\cref{fig:Ncrossv}).} $N$ was always significantly
over-estimated using ABC (\cref{fig:abcpt2}).

In retrospect, it is unreasonable to expect good estimation of $N$, because
adding additional nodes does not change the edge density or overall shape of a
BA network. This can be illustrated by imagining that we add a small number of
nodes to a network after the epidemic simulation has already been completed. It
is possible that none of these new nodes attains a connection to any infected
node. Thus, running the simulation again on the
new, larger network could produce the exact same transmission tree as before.
On the other hand, when $I$ is large relative to $N$, the coalescent dynamics
discussed above also apply. The waiting times until the next infection
increase, resulting in longer coalescence times toward the tips. The relative
accuracy of the nLTT in these situations (\cref{fig:rsquared,fig:Ncrossv})
corroborates this hypothesis, as the nLTT uses only information about the
coalescence times. When all BA parameters were simultaneously estimated with
ABC, $N$ was nearly always over-estimated by approximately a factor of two
(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
bias was our choice of prior distribution. Since the prior for $I$ and $N$ was
jointly uniform on a region where $I \leq N$, more prior weight was assigned to
higher $N$ values.

>>>>

\add{To check that our results were robust to the choice of phylogenetic
method, we constructed phylogenies for two of the datasets (mixed/Spain and
IDU/Estonia) with BEAST~\autocite{drummond2007beast}. We used the TN93 model of
sequence evolution~\autocite{tamura1993estimation}, a strict molecular clock,
and constant population size. The chains were run for $10^7$ steps and sampled
every $10^3$ steps. Maximum clade credibility (MCC) trees were constructed from
the sampled trees after discarding the first 10\% as burn-in.}

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.6\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
    \caption[
        One-dimensional marginal posterior distributions of BA model
        parameters estimated by \textit{netabc} from a simulated transmission
        tree.
    ]{
        One-dimensional marginal posterior distributions of BA model parameters
        estimated by \textit{netabc} from a simulated transmission tree. Dashed
        lines indicate true values, solid lines indicate posterior means, and
        shaded areas show 95\% highest posterior density intervals.
    }
    \label{fig:abc1}
\end{figure}

@@

    joint.I <- d[is.na(fixed), abs(I-true_I)]
    marginal.I <- d[fixed == "N+alpha+m", abs(I-true_I)]

    d <- setDT(collect.data("../../simulations/abc-better-smc/point-estimate/*"))

    base.alpha <- d[nparticle == 1000, abs(alpha - true_alpha)]
    better.alpha <- d[nparticle == 2000, abs(alpha - true_alpha)]

    base.I <- d[nparticle == 1000, abs(I - true_I)]
    better.I <- d[nparticle == 2000, abs(I - true_I)]

    base.N <- d[nparticle == 1000, abs(N - true_N)]
    better.N <- d[nparticle == 2000, abs(N - true_N)]
@

\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
of each of the BA parameters on the same simulated transmission tree. The
estimates of $m$ were apparently unaffected by marginalizing out the other
parameters, corroborating the previous experiments' findings that $m$ is not an
identifiable parameter from scaled tree shapes. Compared to allowing all
parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
    \Sexpr{ round((joint.alpha - marginal.alpha) * 100 / joint.alpha) }\%,
    \Sexpr{ round((joint.I - marginal.I) * 100 / joint.I) }\%, and
    \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
when all other parameters were fixed. \Cref{fig:moresmc} shows the impact of
increasing the number of particles, simulated datasets, and
$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
results of the two simulations were similar, but surprisingly, the results with
higher SMC settings were slightly worse (by
    \Sexpr{ round((better.alpha - base.alpha) * 100 / better.alpha) }\%,
    \Sexpr{ round((better.I - base.I) * 100 / better.I) }\%, and
    \Sexpr{ round((better.N - base.N) * 100 / better.N) }\%
for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
$I$ was slightly closer to the true value of 2000 with the improved settings
    (\Sexpr{ d[nparticle == 2000, round(I.lower.50)] } - 
     \Sexpr{ d[nparticle == 2000, round(I.upper.50)] }, vs.
     \Sexpr{ d[nparticle == 1000, round(I.lower.50)] } - 
     \Sexpr{ d[nparticle == 1000, round(I.upper.50)] } with basic settings).
The estimate of $m$, 3 in both cases, was unaffected by the settings.}

@@

\add{The number of nodes in the network, $N$, exhibited the most variation in
terms of its effect on tree shape. There was almost no measurable difference
between trees simulated under different $N$ values when the number of infected
nodes $I$ was small (\cref{fig:Ncrossv}).} In retrospect, it is unreasonable to
expect good estimation of $N$, in many cases, because adding additional nodes
does not change the edge density or overall shape of a BA network. This can be
illustrated by imagining that we add a small number of nodes to a network after
the epidemic simulation has already been completed. \add{If $I$ is small
relative to $N$, very few of the infected nodes will gain any new neighbours.
Thus, the outcome of a second simulation on the same network will likely be
very similar.} \del{It is possible that none of these new nodes attains a
connection to any infected node. Thus, running the simulation again on the new,
larger network could produce the exact same transmission tree as before.}
On the other hand, when $I$ is large relative to $N$, the coalescent dynamics
discussed above also apply. The waiting times until the next infection
increase, resulting in longer coalescence times toward the tips. The relative
accuracy of the nLTT in these situations (\cref{fig:rsquared,fig:Ncrossv})
corroborates this hypothesis, as the nLTT uses only information about the
coalescence times. When all BA parameters were simultaneously estimated with
ABC, $N$ was nearly always over-estimated by approximately a factor of two
(\cref{fig:abcpt2,tab:abchpd}). One factor which may have contributed to this
bias was our choice of prior distribution. Since the prior for $I$ and $N$ was
jointly uniform on a region where $I \leq N$, more prior weight was assigned to
higher $N$ values. We note also that \del{our accurate estimates of $I$ may
have been influenced by} this prior\del{, which} places more mass on low $I$
values. However, the estimate of $I$ was very high for the HET/Botswana data,
suggesting that a strong enough signal in the data can overcome the prior.
\add{Furthermore, when $I$ was estimated marginally with fixed $N$, the
accuracy of the estimate improved even though there was no longer any extra
prior mass on low $I$ values.}
